---
title: "Data Visualization"
author: "Kiran Favre"
date: "2022-08-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(janitor)
library(here)

library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)

library(gapminder)

library(sf) 
```

### 

```{r}
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(color = "pink")
```

Create a histogram for total lengths of all lizards in data set

```{r}
histogram <- ggplot(data = lizards, aes(x = total_length)) +
  geom_histogram(color = "blue", 
                 fill = "purple", 
                 size = 1, 
                 linetype = "dotted")

#if youre changing something based on a constant, it should not be within aes
```

```{r}
ggplot(lizards, aes( x = total_length,
                     y = weight)) +
  geom_point(
    shape = 7,
    color = "cyan4",
    fill = "pink",
    size = 4,
    alpha = 0.7
  )

#can change shape #1-100
#alpha is transparency: 0 is clear and 1 is opaque
#if you want to change colors based on variables: put in aes
```

```{r}
# things u want constant, put in geom_point. If you want things to change, put it in aes
ggplot(lizards, aes(x = total_length,
                    y = weight)) +
  geom_point(aes(color = common_name,
                 size = total_length,
                 alpha = 0.5)) + 
  theme_dark()

```

```{r}
# use facet wrap to help distinguish patterns across species
ggplot(lizards, aes(x = total_length,
                    y = weight)) +
  geom_point() +
  facet_wrap(~common_name, scales = "free")

#~ is "function of"
```

```{r}
#make a data table with total counts of lizards
# use "new name = n()
# n counts numbers of row

lizard_counts <- lizards |> 
  group_by(common_name) |> 
  summarize(
    counts = n()
  )

lizard_counts <- lizards |> 
  count(common_name)
```

Make a bar plot (column graph) that plots \# of lizards

```{r}
ggplot(lizard_counts, aes( y= fct_reorder(common_name, n),
                            x = n)) +
  geom_col()

#output gives horizontal bar chart in reverse alphabetical order
# to change this, we need to change the class of the species to factor so we can sort it
# use fct_reoder, then need to tell it where the variables are that we want to order it by (which is why we have y = fct_reorder(common_name, n))

```

Create scatter plots faceted by species for total length vs weight. The color and shape of points should differ based on species (common name). Change the theme, and make each faceted panel have flexible axes limits

```{r}
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name, shape = common_name),
             size = 1.5,
             alpha = 0.6) + 
  theme_grey() +
  facet_wrap(~common_name, scales = "free") +
  labs( x = "Total length(mm)",
        y = "weight (g)",
        title = "Jornada Basin lizard sizes",
        subtitle = "All lizards follow the standard weight-length relationship, with Western Whiptails being largest.",
        caption = "Data: Jornada Basin LTER, CCBY",
        )

#using shape only deals with 6 variables, you could 
```

## Afternoon session

### Using ggplot2

#### Scales and color gradients

```{r}
#creating color gradient using two colors and scale_color_gradient
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight),
             size = 3) + 
  scale_color_gradient(low = "purple", high = "red")


# creating color gradient using a vector of colors and scale_color_gradientn
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight),
             size = 3) + 
  scale_color_gradientn(colors = c("magenta", "cyan4",
                                  "darkorchid3", "dodgerblue", "green",
                                  "yellow"))


#diverging color scheme
# creating color gradient using two colors and scale_color_steps
#given an output of color bins
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight),
             size = 3) + 
  scale_color_steps(low = "red", high = "blue")



# creating color gradient using a low, mid and high colors and scale_color_steps2
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = total_length),
             size = 3) + 
  scale_color_steps2(low = "purple",
                     mid = "white",
                     high = "orange",
                     midpoint = 150,
                     breaks = c(50, 70, 150, 180, 220, 280))


#sequential color scheme
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = total_length)) +
               scale_color_stepsn(colors = c("orange",
                                             "red", "purple"),
             breaks = seq(from = 0, 
                          to = 300,
                          by = 60))
```

#### Palettes using paletteer packages

Make a horizontal box plot with common name on the y axis, total length on the x axis, with boxplot fill changing based on the common name

```{r}
#box plot
#reordering the y axis, update sci name using mutate, renaming common_name column (override existing) and factor ize common name variable. Then reorders according to median of total length for each different species 
lizards_median <- lizards |> 
  mutate(common_name = fct_reorder(common_name, total_length,
                                    .fun = median))


ggplot(lizards, aes(y = common_name,
                    x = total_length)) + 
  geom_boxplot(aes(fill = common_name)) +
  labs(x = "Total Length(mm)", 
       y = "Common name")  

```

to check all palette options: `view(palettes_d_names)` in console

```{r}
ggplot(data = lizards_median, aes(y = common_name,
                                  x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) +
  scale_fill_paletteer_d(palette = "awtools"::"a_palette") + 
  theme_minimal()
```

#### Changing elements of a graph using `theme`

```{r}
#element_blank removes that element, so here we will remove y grids
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  theme(panel.grid.major.x = element_line(color = "red"),
        panel.grid.minor.y = element_blank(),
        axis.title.x = element_text(color = "purple", size = 10),
        axis.text.y = element_text(color = "green"),
        text = element_text(size = 18),
        panel.background = element_rect(color = "purple", fill = "yellow")) + 
  annotate("text", x = 300, y = 50, label = "Woooo!", color = "cyan4") +
  geom_vline(xintercept = 250, linetype = "dashed", color = "orange", size = 4)
```

#### Repulsive labels with `ggrepel`

```{r}
wws_lizards <- lizards |> 
  filter(common_name == "western whiptail",
         site == "sand")

ggplot(wws_lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  geom_text_repel(aes(label = toe_num)) #labeling by toe tage number, keeps the value and point itself separate!
```

```{r}
gapminder |>
  filter(year == 2002,
         continent == "Europe") |> 
  ggplot(aes(x = gdpPercap, 
             y = lifeExp)) +
  geom_point() +
  geom_text_repel(aes(label = country), size = 3)
```

#### `gghighlight` highlights based on condition set

```{r}
ggplot(lizards, aes(x = total_length, y = weight)) +
  geom_point(aes()) +
  gghighlight(weight > 30,label_key = toe_num) #anything with weight greater than 30 will be highlighted and labeled with their toe-number
```

### Map making

```{r}
jornada_veg <- read_sf(here("data_raw", "spatial_vegetation", "doc.kml"))

ggplot(data = jornada_veg) +
  geom_sf(aes(fill = Name)) +
  theme_minimal() +
  scale_fill_paletteer_d(palette = "ggsci::amber_material")
```

# Day 10: data viz cont

## Make a heatmap!

```{r}
# first line is storing data in a new column to override date to store in ISO8601
#counting by year and common name
lizard_counts <- lizards |> 
  mutate(date = lubridate::mdy(date)) |> 
  count(year = lubridate::year(date), common_name) |> 
  drop_na()
```

time to make our heatmap!

```{r}
#scale_fill_gradient_n lets u tell R how many colors u want
ggplot(data = lizard_counts, aes(x = year, y = common_name)) +
  geom_tile(aes( fill = n), show.legend = FALSE) +
  geom_text(aes(label = n), color = "white", size = 3) +
  scale_fill_gradientn(colors = c("navy", "red", "orange")) +
  theme_minimal()
```

## Make a beeswarm plot

these give ideas about distribution

more precise than jitter plots

```{r}
whiptails <- lizards |> 
  filter(common_name == "western whiptail") |> 
  drop_na(total_length, weight) #Rows will be dropped that have total length or weight = NA
```

```{r}
whiptail_beeswarm <- ggplot(data = whiptails, aes(x = sex, y = weight)) +
  geom_beeswarm() +
  geom_boxplot(fill = NA, size = 0.5)
```

## Marginal plot

-   starting from whiptails, plot total length (x-axis) versus weight (y axis), with point color dependent on sex

```{r}
whiptail_plot <- ggplot(data = whiptails, aes(x = total_length, y = weight)) +
  geom_point(aes(color = sex)) +
  theme(legend.position = "bottom")


ggMarginal(whiptail_plot, type = "boxplot", groupColour = TRUE)
```

## Patchwork package to create compounded figs

```{r}
((whiptail_beeswarm + whiptail_plot) / whiptail_beeswarm) / 
  whiptail_plot & theme_dark()
# gives programatic control over figures by using operations + - / *
# using "& ____" will attach whatever, theme or sumthin, to all of the plots
```
